---
title:  "A review of statistical computing with TMB"
author: "Andrea Havron<br>NOAA Fisheries, OST"
output: html_document
runtime: shiny
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(mvtnorm)
library(plotly)
#remotes::install_github("AckerDWM/gg3D")
library(gg3D)
```

# ML Inference

What is the likelihood of getting heads 30 times in 100 coin flips?


1. Specify the model <br><br>
$y ~ \sim Binomial(n, p)$


2. Calculate the likelihood<br><br>
$L(p; n, y) = \frac{n!}{y!(n-y)!}p^y(1-p)^{n-y}$


$L(p; n = 100, y = 30)$
```{r, eval=TRUE, out.width = '80%', echo = FALSE}
curve(dbinom(30,100,x),0,0.75, ylab = 'L(p)', xlab = 'p')
```


3. Calculate the negative log-likelihood<br><br>
$-\ell(p; n, y) = -[ln\big(\frac{n!}{y!(n-y)!}\big) + yln(p)$<br>
     $$+ (n-y)ln(1-p)]$$

$-ln\big[L(p; n = 100, y = 30)\big]$
```{r, eval=TRUE, out.width = '80%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
```

4. Calculate the derivative w.r.t. $p$<br><br>
$\frac{d(\ell(p; n, y))}{dp} = \frac{y}{p}- \frac{n-y}{1-p}$



$-ln\big[L(p; n = 100, y = 30)\big]$
```{r, eval=TRUE, out.width = '80%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
```

5. Set to 0 and solve for MLE<br>
$0 = \frac{y}{p}- \frac{n-y}{1-p}$ <br>
$E[p] = \frac{y}{n}$<br>
$E[y] = np$


$-ln\big[L(p; N = 100, y = 30)\big]$
```{r, eval=TRUE, out.width = '70%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
nll <- -dbinom(30,100,.3,TRUE)
segments(0.1, nll, 0.5, nll, col='red', lwd=2)
```
<br>
$\hat{p} = \frac{30}{100} = 0.3$


6. Approximate Var[p] using the second derivative<br>
$-\frac{y}{p^2} - \frac{(n-y)}{(1-p)^2}$<br>
$-\frac{np}{p^2} - \frac{(n-np)}{(1-p)^2}$<br>
$-\frac{n}{p} - \frac{n(1-p)}{1-p}$<br>
$l''(p) = -\frac{n(1-p)}{p}$<br>
$Var[p] = \frac{p}{n(1-p)}$


$Var[p] \approx -\frac{1}{l''(p)}$<br>
$SE[p] \approx \sqrt{ \frac{.3}{100(.7)}} \approx 0.065$<br>
```{r, eval=TRUE, out.width = '70%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
nll <- -dbinom(30,100,.3,TRUE)
points(0.3, nll, col = 'red')
confint <- .3 + c(-1,1)*2*sqrt(.3/70)
segments(confint[1], -log(dbinom(30,100,confint[1])), confint[1], 100, col='red', lwd=2)
segments(confint[2], -log(dbinom(30,100,confint[2])), confint[2], 100, col='red', lwd=2)
text(.3, 20, "95% confint", col='red', cex = 2)
```

# Multivariate asymptotics

* For N-d models, the curvature is represented by a NxN **Hessian** matrix of 2nd partial derivatives
* Inverting the negative Hessian gives us a covariance matrix

\begin{align}
(\mathbb{H}_{f})_{i,j} &= \frac{\partial^2f}{\partial \theta_{i}, \partial x\theta_{j}} = \frac{-1}{Var(\Theta)}
\end{align}

```{r mvnorm, context="render", echo=FALSE, warning = FALSE, message=FALSE}

  library(shiny)
  cor2cov <- function(R,sd2){
    S <- c(sd2, sd2)
    diag(S) %*% R %*% diag(S)
  }
  
  shiny::sliderInput("var", "Variance:", min = 0.1, max = 5, value = 1)

  
  shiny::renderPlot({
    C <- cbind(c(1, .4), c(.4, 1))
    Sigma <- cor2cov(C, input$var)
    x <- seq(-3, 3,  .5)
    y <- x
    z <- matrix(0,length(x),length(y))
    for(i in 1:length(x)){
      for(j in 1:length(y)){
        z[i,j] <- dmvnorm(c(x[i],y[j]), sigma = Sigma)
      }
    }
  persp(x, y, -log(z), zlim = range(0,12))
  })
 
  # df <- as.data.frame(expand.grid(x = seq(-3,3,.5), y = seq(-3,3,.5)))
  # df$z <- 0
  # for(i in 1:nrow(df)){
  #   df$z[i] <-   dmvnorm(c(df$x[i], df$y[i]), sigma = Sigma)
  # }
  # df$z = apply(df, 1, function(t) dmvnorm(c(t[1], t[2]), sigma = Sigma))
  # for(i in 1:length(x.seq)){
  #     df[i,] <- dmvnorm(c(x.seq[i],y.seq[j]), sigma = Sigma)
  #   }
  # # }
  # ggplot(df, aes(x=x, y=x, z=z)) + 
  #   theme_void() + 
  #   gg3D::axes_3D()+
  # stat_wireframe(alpha=.5) +
  # stat_3D(aes(color=z), alpha=.5) +
  # theme_void() +
  # theme(legend.position = "none") +
  # scale_color_gradientn(colors=plot3D::jet2.col()) +
  # labs_3D(hjust=c(0,1,1), vjust=c(1, 1, -0.2), angle=c(0, 0, 90))
  
 # plot.ly(x=df$x, y =)
 
 # print(round(-solve(Sigma),3))

```


